---
title: "Fitting a distributional model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting a distributional model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(statmodels7)
set.seed(1)
```

An ordinary regression models the mean and treats everything else about the
distribution as a nuisance held constant. A distributional model gives **every
parameter its own equation**, so the scale, the shape and a zero-inflation
probability may each depend on covariates, and each may be a smooth, a random
effect or a penalized block.

`statmod()` reads one formula carrying all of them, the equations separated by
a bar, and fits it. This vignette walks through a fit and the questions a reader
asks of one.

```{r}
n <- 300
da <- data.frame(x = sort(runif(n, -3, 3)))
da$y <- rnorm(n, 1 + 0.8 * da$x, exp(-0.6 + 0.35 * da$x))
```

The scale of these data grows with `x`, from about $0.19$ at $x = -3$ to about
$1.57$ at $x = 3$.

## A first fit

The first equation belongs to the distribution's first parameter, so a formula
with no bar models the mean and leaves the rest as intercepts:

```{r}
f1 <- statmod(y ~ x, gaussian1_distrib(), da)
coef(f1)
```

`coef()` answers with **one compartment per parameter**, because the
coefficients of the mean and of the scale are different quantities on different
scales and a single flat vector would run them together. Where the model is an
ordinary regression, it is one:

```{r}
all.equal(unname(coef(f1)$mu), unname(coef(lm(y ~ x, da))), tolerance = 1e-6)
```

## Modeling the scale

Name a parameter on the left of its own equation, after a bar:

```{r}
f2 <- statmod(y ~ x | sigma ~ x, gaussian1_distrib(), da)
lapply(coef(f2), round, 3)
```

against a truth of $\mu = 1 + 0.8x$ and $\log\sigma = -0.6 + 0.35x$. The scale's
equation is on the **log scale**, which is that parameter's own link: an
equation is fitted on the scale that keeps its parameter inside its domain, so
no combination of covariates can produce a negative standard deviation.

Comparing the two fits shows what the second equation bought:

```{r}
c(constant_scale = loglik(f1), modeled_scale = loglik(f2))
```

## Terms

Any `modelterms7` term goes in any equation. A smooth carries a penalty, and its
smoothing parameter is estimated, not set:

```{r}
db <- data.frame(z = sort(runif(n, -3, 3)),
                 g = factor(rep(1:20, length.out = n)))
db$y <- rnorm(n, 1 + sin(1.4 * db$z) + rnorm(20, sd = 0.6)[db$g], 0.4)

f3 <- statmod(y ~ s(z, k = 8), gaussian1_distrib(), db)
f3@edf
```

The smooth carries 7 coefficients and **spends** about 5.3 of them: that is what
the penalty buys, and it is the number a criterion should count, and never the
7. `hyper()` reports the hyperparameters and who chose each:

```{r}
hyper(f3)
```

A random effect is written the same way and is the same machinery, its penalty
being a Gaussian prior over the group effects:

```{r}
f4 <- statmod(y ~ s(z, k = 8) + random(~ 1 | g), gaussian1_distrib(), db)
f4@edf
hyper(f4)
```

20 group effects spend about 18.6 degrees of freedom here, and the estimated
prior scale is close to the 0.6 the groups were drawn at. A random intercept is
a ridge penalty on an indicator block, which is why it needs no separate
machinery.

## Who chooses the hyperparameters

Every hyperparameter is estimated unless the term holds it. Which criterion does
the choosing is `outer_criterion`:

* `reml()` and `ml()` are marginal criteria: they integrate out what the penalty
  shrinks and are differentiable, so the search uses an exact gradient and, where
  it exists, an exact Hessian. `reml()` is the default.
* `aic()`, `bic()` and `cv()` score an estimate of prediction error instead, and
  are what a penalty with a **kink** needs: a marginal criterion expands around
  the penalized mode, and for a lasso that mode sits *at* the kink, where the
  second derivative it wants does not exist.

A term holds a hyperparameter by naming it, and `hyper()` then says so in its
`held` column.

## Reading a fit

```{r}
summary(f3)
```

Three things in that output are worth naming. The log-likelihood is the
**conditional** one, read at the fitted coefficients, and the effective degrees
of freedom go with it: `cAIC` and `cBIC` are built from that pair, and comparing
a conditional log-likelihood against a marginal one from another package is
comparing two different quantities. Every interval is built on the scale that
keeps its quantity in its own set and mapped back, so a scale's interval cannot
contain a negative number. And the last line is the certificate.

## The certificate

The optimizer's convergence flag answers whether a stopping rule fired.
`statmod_certificate()` answers a different question: is the point reported the
one the model asks for?

```{r}
cert <- statmod_certificate(f4)
c(state = cert$state, mode_error = signif(cert$mode_error, 3))
```

`mode_error` is how far above its own penalized mode the inner fit stopped, in
log-likelihood units, so one limit serves every model whatever its scale. The
`state` is `converged`, `boundary` where a hyperparameter has run to an edge and
the criterion is genuinely flat there, or `not converged`.

## Prediction, and the ordinary generics

`predict()` answers per parameter, with a standard error on request:

```{r}
nd <- data.frame(z = c(-1, 0, 1),
                 g = factor(c(1, 2, 3), levels = levels(db$g)))
predict(f4, newdata = nd, se = TRUE)$mu
```

Prediction **reapplies** each term's blueprint rather than rebuilding it, so the
knots of a smooth and the levels of a factor are the ones the model was fitted
with. The identity that pins it is that predicting on rows the model saw returns
the fitted values there.

The `stats` generics behave as they read:

```{r}
c(nobs = nobs(f4), df.residual = df.residual(f4))
length(sigma(f4))
```

`sigma()` returns a **vector**, because the whole point of the layer is that a
scale may be modeled: a single residual standard deviation exists only where
that equation is an intercept, and returning its first value everywhere else
would answer a different question. `df.residual()` subtracts the *effective*
count, so it is not an integer.

Three generics signal an error instead of guessing. `terms()` would have to
choose one equation's terms of several; `model.frame()` has no frame to return,
a fit keeping each term's blueprint and never the data; and `anova()` has no
null distribution to test against when the hyperparameters were chosen from the
same data.

## Simulating from a model

`rstatmod()` draws from a model at coefficients you supply, as a simulation
study needs:

```{r}
sim <- rstatmod(y ~ x, gaussian1_distrib(), da,
                par = list(mu = c(1, 0.8), sigma = -0.6))
names(sim)
head(sim$data, 3)
```

It returns the truth **beside** the data and never as an attribute of it: an
attribute survives a row subset without being subset itself, so
`attr(sim[1:10, ], "theta")` would still hold every row's parameters and be
silently misaligned, which for a simulation study is the worse failure.

## Summary

* One formula, the equations separated by a bar, one per distribution
  parameter; an equation omitted is an intercept.
* Each equation is fitted on its parameter's own link scale, so no combination
  of covariates leaves the parameter's domain.
* `coef()`, `predict()` and `summary()` answer with one compartment per
  parameter.
* Any `modelterms7` term goes in any equation, and its penalty travels with it.
* Hyperparameters are estimated unless a term holds one: `reml()` and `ml()`
  where the penalty is twice differentiable, `aic()`, `bic()` or `cv()` where it
  has a kink.
* The log-likelihood reported is the conditional one, and `cAIC` and `cBIC` go
  with the effective degrees of freedom, never a parameter count.
* `statmod_certificate()` asks whether the point is the model's mode, which the
  optimizer's flag does not.
